---
title: "Mini Data-Analysis Deliverable 1"
author: Nicola Rammell
output: github_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

```         
install.packages("devtools")
devtools::install_github("UBC-MDS/datateachr")
```

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3.  Make a repository in the <https://github.com/stat545ubc-2023> Organization. You can do this by following the steps found on canvas in the entry called [MDA: Create a repository](https://canvas.ubc.ca/courses/126199/pages/mda-create-a-repository). One completed, your repository should automatically be listed as part of the stat545ubc-2023 Organization.

# Instructions

## For Both Milestones

-   Each milestone has explicit tasks. Tasks that are more challenging will often be allocated more points.

-   Each milestone will be also graded for reproducibility, cleanliness, and coherence of the overall Github submission.

-   While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

-   The milestones will be equally weighted.

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 36 points: 30 for your analysis, and 6 for overall reproducibility, cleanliness, and coherence of the Github submission.

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other words, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 **(1 point)** Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: vancouver_trees\
2: flow_sample\
3: building_permits\
4: cancer_sample

<!----------------------------------------------------------------------------->

1.2 **(6 points)** One way to narrow down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->

## Explore vancouver_trees

```{r}
# Use dplyr::glimpse to explore vancouver_trees 
dplyr::glimpse(vancouver_trees) # dataset has 20 columns (variables) and 146,611 rows
                                # most columns contain character data (chr)
class(vancouver_trees)          # dataset is a tibble (tbl_df)
```

The `vancouver_trees` dataset is a tibble with 20 columns and 146,611 rows. Most columns contain character data.

## Explore flow_sample

```{r}
# Use dplyr::glimpse to explore flow_sample
dplyr::glimpse(flow_sample) # dataset has 7 columns (variables) and 218 rows
                            # most columns contain numeric data (dbl)
class(flow_sample)          # dataset is a tibble (tbl_df)
```

The `flow_sample` dataset is a tibble with 7 columns and 218 rows. Most columns contain numeric data.

## Explore building_permits

```{r}
# Use dplyr::glimpse to explore building_permits 
dplyr::glimpse(building_permits) # dataset has 14 columns (variables) and 20,680 rows
                                 # most columns contain character data (chr)
class(building_permits)          # dataset is a tibble (spec_tbl_df)
```

The `building_permits` dataset is a tibble with 14 columns and 20,680 rows. Most columns contain character data.

## Explore cancer_sample

```{r}
# Use dplyr::glimpse to explore cancer_sample 
dplyr::glimpse(cancer_sample) # dataset has 32 columns (variables) and 569 rows
                              # most columns contain numeric data (dbl)
class(cancer_sample)          # dataset is a tibble (spec_tbl_df)
```

The `cancer_sample` dataset is a tibble with 32 columns and 569 rows. Most columns contain numeric data.

<!----------------------------------------------------------------------------->

1.3 **(1 point)** Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 1. What lead you to choose this one? Briefly explain your choice below.

<!-------------------------- Start your work below ---------------------------->

I have decided to use the `vancouver_trees` dataset for my project. Of the four datasets that I explored, this dataset had the most observations, with 146,611 rows of data. This dataset also contains numerous variables of interest, with a mixture of character data (such as species names and street names) and numeric data (such as tree diameter). Overall, I chose this dataset because of its large size, relative completeness, and because I am ultimately interested in learning more about the trees of Vancouver. Specifically, I am interested in how tree biodiversity and species identity vary over space and time.

<!----------------------------------------------------------------------------->

1.4 **(2 points)** Time for a final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with your dataset. Note it down below.

<!-------------------------- Start your work below ---------------------------->

**Research Question:** Do tree communities in Vancouver differ by neighborhood?

<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 **(12 points)** Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*.

Make sure that you're using dplyr and ggplot2 rather than base R for this task. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 **(4 points)** For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->

## Introduction

For Task 2, I will complete Exercises 3, 2, 1, and 8. These are completed out of order because I decided to assess missing data first, create a new variable second, plot the distribution of this variable third, and lastly, create a density plot of another variable. All exercises are completed with the `vancouver_trees` dataset from `datateachr` using the `dplyr` and `ggplot2` packages. 

### Load libraries and data

```{r}
# Load libraries 
library(devtools)                                # package development tools
# devtools::install_github("UBC-MDS/datateachr") # install datateachr from GitHub
library(datateachr)                              # package containing the dataset
library(tidyverse)                               # set of packages for tidy data
library(ggplot2)                                 # package for data visualization
library(ggridges)                                # package for ridgeline plots

# Load and view dataset
trees <- datateachr::vancouver_trees             # create object for the dataset
dplyr::glimpse(trees)                            # explore tibble characteristics
# summary(trees)                                 # to see variable characteristics/NAs
View(trees)                                      # see complete tibble in new window
```

## Exercise 3: Investigate how many missing values there are per variable. Can you find a way to plot this?

**Explanation**: I will begin with Exercise 3 because the first way I wish to explore the dataset is to investigate the number of NA values for each of the columns. With 146,611 rows of data, scrolling through the observations and noting NA frequency is not feasible. Therefore, it makes sense to calculate the number of NAs using code and plot the proportion of NA cells in order to visualize how much of the data is missing. The number of NAs can be viewed using the base::summary function as seen above, but here we will use `dplyr` to achieve our desired result. 

```{r}
# Count number of NAs for each variable and calculate proportion of missing data 
na <- trees %>%
  dplyr::summarise(across(everything(), list(~ sum(is.na(.))))) %>%  # count up NAs
  dplyr::rename_with(~str_remove(., '_1')) %>%                       # remove _1 suffix
  tidyr::pivot_longer(cols = everything(),                           # pivot dataframe
               names_to = "variable_name", 
               values_to = "number_na") %>%                   # number of NAs to a column
  dplyr::mutate(proportion_na = number_na/(nrow(trees)))      # add proportion NAs column
print(na)
```

```{r}
# Next, I will plot the proportion of missing data for each variable 
ggplot(na, aes(x = variable_name, y = proportion_na)) +     
 geom_bar(stat = "identity", fill = "aquamarine3") +
 coord_flip() +
 theme_classic() +
 xlab("Variable") + ylab("Proportion of NA values")
```

**Summary**: Plotting the proportion of NA values for each variable is an effective way to visualize how much data is missing from this dataset. Many variables do not have any missing values, while `plant area` has little missing data (\~1%), `latitude` and `longitude` have an intermediate amount of missing data (\~15%), and `date planted` and `cultivar name` have a substantial amount of missing data (\~50%). If we plan on using the latter two variables for an analysis, we'd need to be cautious since so much data is missing.

**Note**: Beyond using the `is.na` argument, further quality control should be done while assessing missing values from the dataset. For example, in my initial viewing of `summary(trees)` when reading in the data (as seen above), I noticed that the `diameter` variable ranges from 0.00 to 435.00. Let's take a closer look using the `dplyr::count` function. 

```{r}
# Count number of unique diameter values
trees %>%
  count(diameter) 
```
While most observations have a realistic diameter value, some do not. In fact, for 92 trees, a diameter of 0 was recorded. And, there are some instances of unreasonably large diameters, such as the maximum recorded diameter of 435.00 inches. Since these values appear to be measurement or data entry errors, and not NAs, I will leave Exercise 3 as-is. However, this observation is important in considering how the data should be filtered for validity before future analyses. 

## Exercise 2: Create a new variable based on other variables in your data.

**Explanation**: Now that I am aware of the missing data in this dataset, I will move on to some exploratory visualization of the data. However, I will first create a new variable that I am interested in exploring: `year planted`. I will create this variable based on the `date planted` variable, by splitting up the date into multiple columns.

```{r}
# Create new variables (year, month, day) based off the date column in the dataset
trees2 <- trees %>%
  dplyr::mutate(year_planted = lubridate::year(date_planted),   # create year column
                month_planted = lubridate::month(date_planted), # create month column 
                day_planted = lubridate::day(date_planted))     # create day column
View(trees2) # check that 3 new columns have been created 
```

**Summary**: I used dplyr::mutate to create new columns with the `lubridate` package in order to parse dates. The `trees2` dataset now has 3 new columns; I will be using the new `year` column in the next exercise.

## Exercise 1: Plot the distribution of a numeric variable.

**Explanation**: I am interested in plotting the distribution of `year planted`. I have chosen to do this because `year planted` is a numeric variable and I am interested in exploring when most trees in Vancouver were planted. I will plot the distribution using a histogram.

```{r}
# Check that year_planted is a numeric variable 
dplyr::glimpse(trees2)   # using dplyr::glimpse, I see that is it (dbl)
# str(trees2)            # alternatively, could check using str (num)
```

```{r}
# Plot the distribution of year_planted using a histogram 
ggplot(trees2, aes(year_planted)) +
   geom_histogram(aes(y = after_stat(count)),  # number of points in bin
                  fill = "springgreen4", 
                  bins = 30,                   # the number of bins
                  na.rm = TRUE) +              # removing the 76548 rows with NAs
   theme_classic() +
   xlab("Year") + ylab("Number of trees planted") 
```

**Summary**: The above plot shows the distribution of the variable `year planted`. We can see that from the beginning of the dataset in 1989, the number of trees that were planted in a given year initially increased over time. The number of trees planted seems to peak in the mid-late 1990s, after which the number of trees planted plateaus until the early 2010s, after which point the number of trees planted in a given year declines.

## Exercise 8: Use a density plot to explore any of your variables (that are suitable for this type of plot).

**Explanation**: Finally, I am interested in exploring tree `diameter` of tree species. In order to do this, I will create a ridgeline plot using the `ggridges` package. I will plot the `diameter` data for a subset of the data that I am particularly intested in, since the full dataset contains 634 species belonging to 97 genera!

```{r}
# First, let's confirm how many observations per species there are 
trees2 %>%
  count(common_name) # there are 634 species in this dataset 
```

```{r}
# Next, let's confirm how many observations per genus there are 
trees2 %>%
  count(genus_name) # there are 97 genera in this dataset
```

I am interested in exploring tree diameter for *Betula* (aka Birch), a genus which is commonly found in my study system!

```{r}
# What is the range of trunk diameters for Betula species? 
trees2 %>% 
   dplyr::filter(genus_name == "BETULA") %>% 
   dplyr::select(diameter) %>%
   summary() # values range from 2 to 48 inches (DBH); values seem legitimate 
```
Note that units were not provided in the dataset, but are found in the [metadata](https://opendata.vancouver.ca/explore/dataset/street-trees/information/?disjunctive.species_name&disjunctive.common_name&disjunctive.on_street&disjunctive.neighbourhood_name) from the City of Vancouver. Diameter is measured as diameter at breast height (DBH): diameter in inches at a height of 1.3 meters. 

Finally, let's make the density plot for trunk diameter of *Betula* species. 

```{r}
# Create density (ridgeline) plot for Betula using ggridges 
trees2 %>% 
   filter(genus_name == "BETULA") %>% 
   ggplot(aes(colour = common_name, 
              fill = common_name, 
              alpha = 0.5,  # defines transparency
              x = diameter, 
              y = common_name)) +
   ggridges::geom_density_ridges() + 
   xlim(0, 50) +            # define x-axis as 0 to 50, based on the known range
   theme_classic() +
   theme(legend.position = "none") +
   xlab("Diameter at breast height (inches)") + ylab("Common name")
```

**Summary**: The Exercise 8 plot shows kernel density estimates of diameter for different species in the *Betula* genus. According to the plot, species such as River Birch (*Betula nigra*) and Gray Birch (*Betula populifolia*) had large diameters at the time of planting, whereas species such as Young's Weeping Birch (*Betula pendula*) and While Himalayan Birch (*Betula jacquemontii*) had small diameters at the time of planting.

<!----------------------------------------------------------------------------->

# Task 3: Choose research questions

**(4 points)** So far, you have chosen a dataset and gotten familiar with it through exploring the data. You have also brainstormed one research question that interested you (Task 1.4). Now it's time to pick 4 research questions that you would like to explore in Milestone 2! Write the 4 questions and any additional comments below.

<!--- *****START HERE***** --->

**Research Questions:**

1.  What is the distribution of common and rare tree species in Vancouver, BC?\
2.  Does tree community composition differ by neighborhood?\
3.  How does tree species richness vary by neighborhood?\
4.  Have the species identity of planted trees changed over time?

<!----------------------------->

# Overall reproducibility/Cleanliness/Coherence Checklist

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. An example of a major continuity error is having a data set listed for Task 3 that is not part of one of the data sets listed in Task 1.

## Error-free code (3 points)

For full marks, all code in the document should run without error. 1 point deduction if most code runs without error, and 2 points deduction if more than 50% of the code throws an error.

## Main README (1 point)

There should be a file named `README.md` at the top level of your repository. Its contents should automatically appear when you visit the repository on GitHub.

Minimum contents of the README file:

-   In a sentence or two, explains what this repository is, so that future-you or someone else stumbling on your repository can be oriented to the repository.
-   In a sentence or two (or more??), briefly explains how to engage with the repository. You can assume the person reading knows the material from STAT 545A. Basically, if a visitor to your repository wants to explore your project, what should they know?

Once you get in the habit of making README files, and seeing more README files in other projects, you'll wonder how you ever got by without them! They are tremendously helpful.

## Output (1 point)

All output is readable, recent and relevant:

-   All Rmd files have been `knit`ted to their output md files.
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

(0.5 point deduction if any of the above criteria are not met. 1 point deduction if most or all of the above criteria are not met.)

Our recommendation: right before submission, delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant. Then, after your final commit and push to Github, CHECK on Github to make sure that everything looks the way you intended!

## Tagged release (0.5 points)

You've tagged a release for Milestone 1.

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
