---
title: "Mini Data-Analysis Deliverable 2"
author: "Nicola Rammell"
output:
  github_document:
    toc: true
---

*To complete this milestone, you can edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by:

-   Making summary tables and graphs
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.*

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission.

**Research Questions**: In Milestone 1, you chose two research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

```{r, message = FALSE}
library(datateachr) # <- contains the tree data I picked!
library(tidyverse)  # will load ggplot2, dplyr!
library(tidyr)      # for tidying data
library(forcats)    # for reordering factors 
library(broom)      # for model output objects to tibbles
```

# Task 1: Process and summarize your data

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions.

### 1.1 (1 point)

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->

**1.** What are the common and rare tree genera and species planted in Vancouver, BC?\
**2.** How does tree species richness vary by neighborhood?\
**3.** Does tree community (taxonomic) composition differ by neighborhood?\
**4.** Have the species identity of planted trees changed over time?

<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4.  Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:**

6.  Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7.  Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks:

8.  Create a graph that has at least two geom layers.
9.  Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->

**Please Note:** For the Summarizing Options 1-4, Option 1 and Option 3 require a numerical variable. However, none of my 4 research questions are based on a numerical variable! So, I will complete Option 2 twice (for RQs 1-2) and Option 4 once (for RQ 3). For RQ 4, I will treat the variable `year_planted` as a numerical variable for the purpose of completing Option 3. I will do this because I would like to group `year_planted` into categories of "1980s", 1990s", "2000s", and "2010s" for the purpose of this exercise.

#### Research Question 1: What are the common and rare tree genera and species planted in Vancouver, BC?

**Summarizing:** I will complete Option 2, computing the number of observations for a categorical variable. This helps answer the RQ because first, I will compute the number of observations for the categorical `genus_name` variable to gain insight into which genera are common and which are rare. Next, I will compute the number of each tree species (by `common_name`) planted in Vancouver, BC. In other words, by ordering values from small to large, I will find out which are the most rare and common tree genera and species planted!

```{r}
# View dataset and confirm genus_name and common_name are character variables
View(vancouver_trees)            # view complete df in new window
dplyr::glimpse(vancouver_trees)  # genus_name and common_name = <chr>

# compute number of observations for categorical genus_name variable
task1.2.1.1 <- vancouver_trees %>%
  dplyr::count(genus_name) %>%
  dplyr::arrange(n)
head(task1.2.1.1)  # rare genera include Albizia, Halesia, and Kalopanax 
tail(task1.2.1.1)  # common genera include Fraxinus, Prunus, and Acer

# next, also compute number of observations for each species
task1.2.1.2 <- vancouver_trees %>%
  dplyr::count(common_name) %>%
  dplyr::arrange(n)
head(task1.2.1.2)  # rare species include Alpine Fir, Bitter Cherry, Bosc Pear
tail(task1.2.1.2)  # common species include Kwansan Flowering Cherry, Pissard Plum
```

**Graphing:** I will complete Option 6, making a barplot where one of the axes is logarithmic, and the axis labels are formatted "pretty", as seen by the subscript. Specifically, I will plot log~10~(Number of trees planted) by Genus. This will directly help answer the RQ because it will show the most common genera and the most rare genera! I will also order the bars from most common to most rare to help illustrate this pattern more clearly. As seen by the plot below, *Acer* (Maples) and *Prunus* (includes Cherry trees) were the most commonly planted genera in Vancouver from 1989 - 2019.

```{r fig1-barplot-logarithmic}
# barplot with number of trees planted vs. genus to see rare and common genera
task1.2.1.1 %>% 
  ggplot(aes(x = fct_rev(fct_reorder(genus_name, n)), 
             y = n)) +                                      # order common to rare
  geom_bar(stat = "identity", fill = "darkgreen") +
  scale_y_log10() +                                         # y-axis is log10
  theme_classic() +                                         # classic theme
  theme(axis.text.x = element_text(angle = 90)) +           # x-axis labels vertical
  theme(axis.text.x = element_text(size = 5)) +             # make text smaller
  ylab(expression(log[10] ~ "(number of trees)")) + # make label "pretty"
  xlab("Genus") 
```

#### Research Question 2: How does tree species richness vary by neighborhood?

**Summarizing:** Next, I will complete a variation of Option 2, computing the *unique* number of observations for a categorical variable. This directly helps answer the RQ because I will now compute the number of unique common names in each of the neighborhood categories. In other words, I will take a first look at species richness in each neighborhood!

```{r}
# compute number of (unique spp.) observations for neighborhood variable
task1.2.2 <- vancouver_trees %>% 
  dplyr::group_by(neighbourhood_name) %>% 
  dplyr::summarise(species_richness = n_distinct(common_name)) %>%
  dplyr::arrange(species_richness) %>%
  ungroup()
print(task1.2.2)
```

**Graphing:** I will complete Option 8, by creating a graph that has at least two geom layers. Specifically, I will make a barplot of `species_richness` vs. `neighbourhood_name`, but I will add a second geom layer to overlay text onto the plot. The text will show the specific species richness value for each neighborhood, eliminating the need for text on the y-axis. This will directly help answer the RQ, by clearly showing how tree species richness varies by neighborhood.

```{r fig2-barplot-layered}
# barplot with species richness vs. neighborhood with added geom layer
task1.2.2 %>% 
  ggplot(aes(x = fct_reorder(neighbourhood_name, species_richness), 
           y = species_richness)) +                   # order low to high
  geom_bar(stat = "identity", fill = "pink4") +            
  theme_classic() +                                   # classic theme
  theme(axis.text.x = element_text(angle = 90)) +     # x-axis labels vertical
  theme(axis.text.x = element_text(size = 8)) +       # text size
  ylab("Species richness") +
  xlab("Neighbourhood")  +
  theme(axis.text.y = element_blank(),                # remove y-axis labels
        axis.ticks.y = element_blank()) +             # remove y-axis ticks
  ylim(0, 450) + 
  geom_text(aes(label = species_richness), size = 3,  # add geom layer for text
            position = position_dodge(width = 0.5), vjust = -0.75) 
```

#### Research Question 3: Does tree community composition differ by neighborhood?

**Summarizing:** I will complete Option 4, computing the proportion and counts in each category of one categorical variable across group of another categorical variable. This helps answer the RQ because I will compute the *proportion* that each tree genera makes up in each neighborhood. This will provide a first look into the overall community composition by neighborhood, as guided directly by the research question!

Note: This exercise will first pivot data into the "wide" format, which is not considered tidy data. However, it is generally the standard format to view species composition data, so I will use the `pivot_wider` function, being aware of the limitations of this format. Next, I will use the `pivot_longer` function to create a "tidy" version of the dataframe, as well!

```{r}
# compute genera counts by neighborhood using pivot_wider
task1.2.3.1 <- vancouver_trees %>%
  dplyr::select(neighbourhood_name, genus_name) %>%
  dplyr::group_by(neighbourhood_name, genus_name) %>%
  dplyr::summarise(n = n()) %>%           # adds column with counts of each genus
  ungroup() %>%
  tidyr::pivot_wider(names_from = genus_name, 
                     values_from = n) %>% # pivot so each genus is now a column
  replace(is.na(.), 0) %>%                # replace NAs with zeros
  dplyr::rowwise() %>%                    # execute next operation across each row
  mutate(TOTAL = sum(c_across(2:98)))     # adds a TOTAL (sum) column
print(task1.2.3.1)  
```

The above wide-format tibble shows the *counts* in each genus category by neighborhood category. A "total" column is added at the end, which shows the sums (total count per neighborhood).

```{r}
# next, use pivot_longer to create a tidy version of this dataset
task1.2.3.2 <- task1.2.3.1 %>%
  select(-TOTAL) %>% 
  tidyr::pivot_longer(
    cols = ABIES:VIBURNUM, 
    names_to = "genus", 
    values_to = "counts")
head(task1.2.3.2)
```

The above tibble uses `pivot_longer` to transform the dataframe into a long farmat that is more useful for data visualization. This tibble will be used in the graphing exercise below!

```{r}
# also compute genera proportions by neighborhood (convert counts to proportion)
task1.2.3.3 <- task1.2.3.1 %>%
  dplyr::mutate(across(c(2:99), .fns = ~./TOTAL))  # divide by the sum
head(task1.2.3.3)
```

The above wide-format tibble shows the *proportions* in each genus category by neighborhood category. A "total" column is added at the end, which shows the sums (total = 1).

```{r}
# next, use pivot_longer to create a tidy version of this dataset
task1.2.3.4 <- task1.2.3.3 %>%
  select(-TOTAL) %>% 
  tidyr::pivot_longer(
    cols = ABIES:VIBURNUM, 
    names_to = "genus", 
    values_to = "proportion")
head(task1.2.3.4)
```

Similarly, the above tibble uses `pivot_longer` to transform the dataframe into a long format that is more useful for data visualization. This tibble will also be used in the graphing exercise below.

**Graphing:** I will complete Option 8, by again creating a graph that has at least two geom layers. This time, I will be plotting the counts of each genus making up the planted tree community in each neighborhood. This helps answer the research question because it allows us to visualize how tree community composition (by genera) varies across the neighborhoods. To add a second geom layer, I will again add text to the plot. This time, I will add labels to the largest category, *Acer*, in the stacked barplot.

```{r fig3-barplot-layered}
# barplot with tree counts by genus vs. neighborhood with added geom text layer
task1.2.3.2 %>%
  ggplot(aes(x = neighbourhood_name, y = counts, fill = genus)) +
  geom_col() + 
  theme_classic() +
  coord_flip() +
  theme(axis.text.x = element_text(size = 7)) +       
  ylab("Number of trees planted") +
  xlab("")  +
  theme(legend.position = "none") + 
  geom_text(aes(label = ifelse(genus == "ACER", # if genus = ACER, add text  
                               as.character(counts), "")), 
                position = position_stack(vjust = .5), 
                size = 2.5, 
                colour = "black")
```

For clarity, the legend has been removed from the above plot, since we are not necessarily interested in seeing the identity of all genera per se. Instead, we may be interested in viewing how many *Acer* trees are planted relative to the other genera in each of the neighborhoods. If this is of interest, we could make a plot like the one above that shows the counts for the *Acer* genus to help answer the research question. The above plot shows both the total number of trees planted in each neighborhood, as well as the contribution made by each genus.

However, to answer the question of how community composition differs between the neighborhoods, we may also wish to see a plot based directly on *relative* abundances as seen below. For clarity, the legend has again been removed from the plot.

```{r fig4-barplot-layered}
# barplot with tree proportions by genus vs. neighborhood 
task1.2.3.4  %>%
  ggplot(aes(x = neighbourhood_name, y = proportion, fill = genus)) +
  geom_col() + 
  theme_classic() +
  coord_flip() +
  theme(axis.text.x = element_text(size = 7)) +       
  ylab("Relative abundance") +
  xlab("")  +
  theme(legend.position = "none") 
```

For clarity, the legend has again been removed from the above plot. Abundant genera are *Acer* (salmon), *Prunus* (purple), *Fraxinus* (green), and *Tilia* (pink)

#### Research Question 4: Have the species identity of planted trees changed over time?

**Summarizing:** I will complete Option 3, creating a categorical variable (with 3 or more groups) from the "numerical" (please see the Note above on line 103) `year` variable. This helps answer the RQ because we may be interested in how the species identify of planted trees has changed by comparing one decade to the next, instead of year by year. Therefore, I will calculate a new variable called `decade` with the categories of "1980s", 1990s", "2000s", and "2010s".

```{r}
# create year_planted variable from date_planted variable for this exercise
task1.2.4.0 <- vancouver_trees %>%
  dplyr::mutate(year_planted = lubridate::year(date_planted),   # create year column
                month_planted = lubridate::month(date_planted), # create month column 
                day_planted = lubridate::day(date_planted))     # create day column
head(task1.2.4.0)

# check that year_planted can be considered a "numerical variable" for this exercise
dplyr::glimpse(task1.2.4.0)  # OK, year_planted is <dbl>

# also, check range of values in the year_planted variable
summary(task1.2.4.0$year_planted)  # values range from 1989-2019
```

```{r}
# now, create a new categorical variable from the year_planted variable
task1.2.4.1 <- task1.2.4.0  %>%
  mutate(decade = factor(case_when(year_planted < 1990 ~ "1980s",
                            year_planted < 2000 ~ "1990s",
                            year_planted < 2010 ~ "2000s", 
                            year_planted < 2020 ~ "2010s"), 
                         levels = c('1980s', '1990s', '2000s', '2010s')))
View(task1.2.4.1)              # confirm values all look correct in new window
# dplyr::glimpse(task1.2.4.1)  # note `decade` is factor with 4 levels
head(task1.2.4.1)  
```

**Graphing:** I will complete Option 7, making a graph where it makes sense to customize the alpha transparency. Here, I will graph the number of trees by species (for a specific genus of interest) planted over time. Therefore, I will also be using a one of the new variables I created in the summarizing task above: `year_planted`. For this graph, it makes sense to customize the alpha transparency simply because there are many overlapping lines, and I would like them all to be visible. Without the transparency feature, some lines would be completely overridden by others!

I am interested in the Ash (*Fraxinus*) genus, because of their significance for bird forage, so I will plot data from this group. This helps answer the research question directly because it helps visualize how the species identity of planted trees changes over time; however, note this is only for one particular genus of interest. The answer may be different for different genera.

```{r}
# count up total numbers of trees planted by common name in each year
task1.2.4.2 <- task1.2.4.0 %>%
  dplyr::filter(!is.na(year_planted)) %>% # removes 76548 rows (recall Milestone 1)
  dplyr::group_by(year_planted, genus_name, common_name) %>%
  dplyr::summarise(n = n()) %>%           # adds column with counts of each common name
  ungroup()
head(task1.2.4.2)
```

```{r fig5-lineplot}
# plot number of Ash tree species planted over time
task1.2.4.2 %>%
  filter(genus_name == "FRAXINUS") %>% 
  ggplot(aes(year_planted, n)) +
  geom_line(aes(colour = common_name), alpha = 0.5) +  # adds transparency
  ylab("Number of trees planted") +
  xlab("Year") +
  scale_colour_discrete("") +
  theme_classic() +
  theme(legend.position = "top") + 
  theme(legend.text = element_text(size = 5))  +
  theme(plot.margin = unit(c(0.2, 0.2, 0.2, 0.2), "inches")) 
```

The above plot suggests that yes, the identity of planted Ash trees did change over time. In the early years shown on the plot, there are many Raywood Ash trees planted (shown in pink), whereas in the more recent years shown on the plot, Autumn Applauce and Flowering Ash (shown in Brown and Green, respectively) seen to become the most common Ash trees planted.

<!----------------------------------------------------------------------------->

### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->

Based on the 8 operations completed, I have made progress in answering each of my research questions.

For RQ1, *What are the common and rare tree genera and species planted in Vancouver, BC?*, I was able to quickly view the answers in the form of summary tables. I was also able to produce a plot that showed the frequency of planted genera from most common to most rare.

For RQ2, *How does tree species richness vary by neighborhood?*, I was also able to produce a plot that demonstrated the species richness in each neighborhood. Similarly, for RQ3, *Does tree community (taxonomic) composition differ by neighborhood?*, I produced plots that showed the total (counts) and relative abundance (proportion) of the different genera planted in the different neighborhoods. These were useful exploratory visualizations; however, these operations did not actually answer the questions with any kind of statistical test. I am still unclear on these two research questions.

Lastly, for RQ4, *Have the species identity of planted trees changed over time?*, I was able to delve into this question for one genus of interest and show that likely, yes, the species identity of planted Ash trees did actually change over time. This was an interesting observation, however, again this question was not answered with any certainty since there was no statistical test involved and this was only done for one particular genus of interest.

Overall, I am interested in looking more into the questions regarding how species richness and community composition differ by neighborhood, since these questions still remain the most unclear. Specifically, I would like to refine the research question to ask about the difference between East and West Vancouver neighborhoods, so that the dataset can provide multiple samples per category.

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->

```{r}
# view the original vancouver_trees dataframe and review columns
View(vancouver_trees)
glimpse(vancouver_trees)
```

Overall, yes, my dataset is already tidy. Each column represents its own variable, and each cell has only 1 value, with the exception of the `date_planted` column which is discussed below.

In the numeric `tree_id`, `civic_number`, `height_range_id`, `diameter`, `latitude`, and `longitude` columns, each row is an observation and each cell has only one value. If the data weren't tidy, the `latitude`, and `longitude` data may be included in the same cell, for example.

The character data columns are also tidy: `genus` and `species` are in separate columns, for example. If the data were not tidy, the *Genus species* information may be included all in one cell. This would have made it very difficult to do the above analyses; we would have had to separate the column into two before proceeding!

Tidying of the data could be improved by separating the `date_planted` column into 3 columns: `year_planted`, `month_planted`, and `day_planted`. That way, if we wish to analyse the data by month, irregardless of year for example, this wouldn't require any further data manipulation.

<!----------------------------------------------------------------------------->

### 2.2 (4 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->

To begin with, the `vancouver_trees` dataframe can be considered tidy, as discussed above.

```{r}
# see the tidy dataframe 
head(vancouver_trees)
tail(vancouver_trees)
```

So, I will create an untidy dataframe below. To do this, I will put `latitude` and `longitude` together into 1 column, and `genus_name` and `species_name` together into 1 column. I am doing this because I think this is a common mistake in datasets, and if I am to run into this issue in the future, I would like to know how to separate them into different columns! Since I will need to "tidy" the "untidying" I am doing to this dataset below, this is a useful and practical exercise to try.

```{r}
# create df with latitude + longitude together in 1 column, separated by ", "
task2.2.0 <- vancouver_trees %>% 
  tidyr::unite(coordinates, latitude,longitude, sep = ", ")
View(task2.2.0)    # see coordinates are now in the form: 49.217759, -123.114718
head(task2.2.0)  

# next, create df with genus + species together in 1 column, separated by " "
task2.2.1 <- task2.2.0 %>% 
  tidyr::unite(scientific_name, genus_name,species_name, sep = " ")
View(task2.2.1)  # see names are now in the form: SORBUS ARIA
head(task2.2.1)
```

The above dataframe has a `coordinates` column in the form "latitude, longitude" because I think this is how datasets often have coordinates stored, delimited by a comma. On the other hand, I made the `scientific_name` column in the form "GENUS SPECIES" because we wouldn't expect data to come with any delimiter when it comes to scientific names.

Next, I will tidy the now untidy dataframe below.

```{r}
# see the untidy dataframe
head(task2.2.1)
tail(task2.2.1)

# now, tidy the dataframe!
task2.2.2 <- task2.2.1 %>% 
  tidyr::separate_wider_delim(coordinates, ",", 
                       names = c("latitude", "longitude")) %>%
  tidyr::separate_wider_delim(scientific_name, " ", 
                       names = c("genus", "species", NA, NA, NA), # deals with Xs
                       too_few = "align_start") %>%               # keep 1st two elements
  dplyr::mutate(year_planted = lubridate::year(date_planted),     # create year column
                month_planted = lubridate::month(date_planted),   # create month column 
                day_planted = lubridate::day(date_planted)) %>%   # create day column
  dplyr::select(-date_planted)
head(task2.2.2)
tail(task2.2.2)
```

As seen in the dataframe above, `genus`, `species`, `latitude`, and `longitude` data have all been returned to their own separate columns. Separating and re-uniting the genus and species data actually improved the "tidyness" of the dataset, because in the original dataset there were "Xs" in some of the cells along with the genus or species name. These "X" probably come from crossing cultivar strains that should have been in the cultivar column. By retaining only genus and species, I have tidyied up the scientific name columns. Additionally, I separated the date column into `date_planted`, `month_planted`, and `day_planted` so that each cell contains only one piece of information. The dataframe is now tidy!

<!----------------------------------------------------------------------------->

### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->

Based on my explanations in Task 1.3, I am slightly revising my research questions. The research questions that I am now most intersted in answering are:

**1.** Does planted tree species richness vary between East and West Vancouver neighborhoods? **2.** Does planted tree community composition differ between East and West Vancouver neighborhoods?

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

<!--------------------------- Start your work below --------------------------->

I have decided to select two research questions based closely on my original Research Questions 2-3 from Milestone 1. These two research questions were: *How does tree species richness vary by neighborhood?*; and *Does tree community composition differ by neighborhood?*

As described in Task 1.3, these are the two research questions that I was least able to answer through my various exploratory operations in Task 1.2. Based on the graphs that I created, there does appear to be a difference in species richness and maybe in community composition between the neighborhoods, but the answers to these questions are still not clear.

Since I only have 1 value of species richness per neighborhood in the dataframe that I created to answer this question, I would like to slightly revise both of these questions. Working through the exercises lead me to revise my two research questions to instead be: *Does planted tree species richness vary between East and West Vancouver neighborhoods?*; and *Does planted tree community composition differ between East and West Vancouver neighborhoods?*

I am interested in exploring these research questions further because it was difficult to obtain an answer based on previous exploratory analyses. I am also interested to know whether there is a difference in the City of Vancouver's tree planting in generally more wealthy (West) versus less wealthy neighborhoods (East). My neighborhood is in East Vancouver in Hastings-Sunrise, and I would also like to see how this neighborhood fits into the overall pattern with the rest!

<!----------------------------------------------------------------------------->

Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidying, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.)

<!--------------------------- Start your work below --------------------------->

To answer the first question, *Does planted tree species richness vary between East and West Vancouver neighborhoods?*, I will need to make a dataframe that includes `neighboorhood_name`, `species_richness`, and `location` which will contain data labelled either East or West. I will create this column based off the existing `longitude` column. To do so, I will take the mean longitude value for each neighborhood, and use this to assign each neighborhood to a category of "East" or "West". This will be based on the Vancouver Main Street longitude coordinates of -123.100796.

```{r}
# create a dataframe with the new location column
task2.3.1 <- vancouver_trees %>% 
  dplyr::group_by(neighbourhood_name) %>% 
  dplyr::summarise(species_richness = n_distinct(common_name),  # calculate spp. richness
            mean_longitude = mean(longitude, na.rm = TRUE)) %>% # calculate avg longitude
  dplyr::mutate(mean_longitude = mean_longitude*(-1)) %>%       # make value positive
  dplyr::mutate(location = factor(case_when(mean_longitude < 123.100796 ~ "East",
                                 mean_longitude > 123.100796 ~ "West"), 
                                 levels = c('East', 'West')))   # add East and West column
print(task2.3.1)
```

The above dataframe could now be used to answer my new research question! Before proceeding, I would like to first visualize the new dataframe with species richness data by East and West Vancouver neighborhoods by using a boxplot.

```{r fig6-boxplot}
# plot species richness by East and West side
task2.3.1 %>%
  ggplot(aes(location, species_richness)) +
  geom_boxplot(aes(), fill = "darkgreen") +
  theme_classic() +
  ylab("Species richness") +
  xlab("")
```

By the looks of it, species richness of planted trees is actually higher in East Vancouver neighborhoods. Now that we have this dataframe, we could use it for test if this relationship is actually significant in Task 3!

To answer the other question, *Does planted tree community composition vary between East and West Vancouver neighborhoods?*, I will need a different version of the data. For this question, I can use a version of the data similar to the one I created in Task 1.2: relative abundance of each tree genus by neighborhood. Again, the only difference is that I will need to add the `location` column so that the data can be compared between the two groups.

To save time, I will simply add the `location` column that I just created in this section Task 2.3 to the previous dataframe that I created back in Task 1.2. I can do this because the dimensions are the same (22 rows = 22 neighborhoods) and in both cases the neighborhoods are sorted alphabetically so the new `location` column can be added seamlessly.

```{r}
# view old dataframe from Task 1.2, Research Question 3
print(task1.2.3.3)  # 22 rows alphabetically from ARBUTUS-RIDGE to WEST POINT GREY

# view new dataframe for *new* research question from this Task 2.3
print(task2.3.1)    # 22 rows alphabetically from ARBUTUS-RIDGE to WEST POINT GREY
task2.3.1$location  # this column can be added to old dataframe

# add the location column to the old dataframe to create the new dataframe I need
task2.3.2 <- task1.2.3.3 %>%
  tibble::add_column(location = task2.3.1$location) %>%   # add new location column
  dplyr::relocate(location, .after = neighbourhood_name) %>%  
  dplyr::select(-TOTAL)                                   # remove TOTAL column
print(task2.3.2)
```

The above dataframe could now be used to answer my second new research question! The dataframe is now set up in the format used to run an NMDS or other ordination technique, for example. The two groups (East and West) could then be compared using a multivariate approach like MANOVA. IN addition, I will also make a tidy version of this dataset below:

```{r}
# use pivot_longer to create a tidy version of the dataframe
task2.3.3 <- task2.3.2 %>%
  tidyr::pivot_longer(
    cols = ABIES:VIBURNUM, 
    names_to = "genus", 
    values_to = "proportion")
head(task2.3.3)
```

The above versions of the data will be appropriate to answer my two research questions. I used a variety of functions to create them, including relocate, select, mutate, summarise, and group_by from the `dplyr` package and pivot_longer from the `tidyr` package. 

<!----------------------------------------------------------------------------->

# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**: Does planted tree species richness vary between East and West Vancouver neighborhoods?

**Variable of interest**: Species richness

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

<!-------------------------- Start your work below ---------------------------->

I will fit a linear model to see how well the location ID predicts the species richness of planted trees. The dataframe I am using is the first of the dataframes that I prepared above in Task 2.3. 

Note that since I am comparing two groups, I could equally have used a t.test() to compare the means across the two groups. Here, I have instead chosen to fit a lm, which will do an equivalent job and also compare the two groups, only by setting "locationEast" as the control group for which to compare "locationWest". 

```{r}
model1 <- lm(species_richness ~ location, data = task2.3.1)
model1
summary(model1)
```

<!----------------------------------------------------------------------------->

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.

<!-------------------------- Start your work below ---------------------------->

I will produce a p-value from the above fitted model. This can be done using `broom::tidy` or `broom::glance`, and the specific value I am interested in can also be printed as seen below. 

```{r}
# use broom::tidy to see p.value = 3.006993e-01
task3.2.1 <- broom::tidy(model1)
print(task3.2.1)
          
# alternatively, use broom::glance to see p-value = 0.3006993
task3.2.2 <- broom::glance(model1)
print(task3.2.2)

# or, print just the value I am interested in (p-value) = 0.3006993	
pvalue <- task3.2.1[2, 5]
pvalue
```

The p-value from the above model is 0.3, indicating that neighborhood location is *not* a good predictor of species richness. While there did appear to be some difference in species richness between the East and West locations in the boxplot above, this difference is not significant. This is unsurprising given that the boxplot showed the interquartile ranges are largely overlapping one another. 

<!----------------------------------------------------------------------------->

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

<!-------------------------- Start your work below ---------------------------->





<!----------------------------------------------------------------------------->

## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.

<!-------------------------- Start your work below ---------------------------->

<!----------------------------------------------------------------------------->

# Overall Reproducibility/Cleanliness/Coherence Checklist

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors.

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone.

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files.
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2.

### Attribution

Thanks to Victor Yuan for mostly putting this together.
